Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-04-16 forecastflows.tex
removing some comments before sending
start to incorporate J&J feedback

\title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology } 
\date{\today}
\author{Victor Van der Meersch, J. Regetz, T. J. Davies$^*$ \& EM Wolkovich}
\maketitle
$^*$ Says he is happy to help and give friendly review, but not sure he will reach level of co-author (though Lizzie used his LPI example).

@@

% To do:
% Fix the abstract so it matches the paper better (I think mainly the last few sentences need work, or we just need to say more about data sharing in the text)
% Add in Figures
% Delete out unnecessary comments so easier for co-authors to read?

\begin{abstract}
Increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts. Relevant datasets to meet these demands have also increased in size and complexity, including in their geographical, temporal and phylogenetic scales. While new research often suggests that accounting for these complexities variously increases, removes or otherwise alters major trends, we argue that the fundamental approach to model fitting in ecology makes it impossible to evaluate and compare models. These problems stem in part from continuing gaps between statistical workflows---where the data processing and model development are often addressed separately from the ecological question and aim---and scientific workflows, where all steps are integrated. Yet, as ecologists become increasingly computational the opportunity to close this gap has never been greater. We outline how increased data simulation at multiple steps in the scientific workflow could revolutionize our understanding of ecological systems, yielding new insights for both trend estimation and forecasting.
%Combining these changes with more open model and data sharing---and developing new efforts to race the same data---could be transformative for ecological forecasting. 
A shift toward universal training in a more robust model building could bridge the gap between process-based and statistical approaches and be transformative for ecological modeling.
\end{abstract}

@@

Anthropogenic drivers are reshaping natural systems \citep{Diaz2019}. Impacts are projected to increase in coming decades, as climate change accelerates biodiversity loss, altering ecosystem services and human well-being \citep{IPBES2019}. Implementing sustainable policies to mitigate these impacts is thus a global priority, but designing the best policies requires estimating and understanding biodiversity and ecosystem trends to date alongside the skill to forecast future dynamics. % These requirements in turn have created a growing need for more global ecological data, and the types of models that can robustly handle such data.

@@

Workflows that fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and degeneracies, before estimating its parameters and making projections, could reduce many of these problems. In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data. Towards this aim,  we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.

@@

Quantitative science relies on a model-based framework, to confront hypothesis with data \citep{Chamberlin:1965cd}. In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses. But this idealized method often does not apply to the reality of ecological research. Many important questions cannot be addressed through controlled experiments and replications. Instead, we must often rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}.

This reality should drive researchers to use more robust and coherent methods, but the current workflows combined with the challenges ecologists are facing instead may lead to persistent problems. 
Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way that makes feedbacks to highlight uncertainty and related limitations in the model and/or data rare (Fig. \ref{fig:modeldata}). For forecasting, researchers have focused on making predictions with increasingly complex models, increasingly obscuring the steps underlying model building and parameterization (Fig. \ref{fig:modeldata}). Today, researchers often calibrate different parts of the model separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole. Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow. 
% mention the disconnection between trends and forecasts?

\begin{figure}[h]
	\centering
	\includegraphics{figures/modeldataspaces}
	\caption{} 
	\label{fig:modeldata}
\end{figure}

We argue a workflow that moves along the data-model space in a coherent sequence of steps with repeated data simulation (Fig. \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
The first step of this workflow is to define an explicit research question and formulate hypotheses (Fig. \ref{fig:workflowy}). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This step should help focus on the mechanisms that could generate the data we observe, including the observational error. Naturally, this leads to the development of a model---an ensemble of mathematical equations that encapsulates our knowledge and is designed to answer our research question. The general idea is to start with a relatively simple model that we could refine later. At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values.

With a model in place the next step focuses on testing and understanding it via data simulation. `Fake' or `test' data are generated directly from the model by fixing parameters to some values (which is straightforward if the parameters are interpretable) and from fake predictor data (Fig. \ref{fig:workflowy}). This simulated data should reflect the full model assumptions, and could begin to include complexities in our data structure and biases, which may in turn lead to adjusting the model. For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step. We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values.

Once we are confident about our model structure, we can incorporate real data. This way, we obtain parameter estimates constrained by observations. Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question. This could lead us to either simplify our research question or---ideally---launch new data collection efforts. This leads to the second data simulation step, this time using our fitted model parameters to generate predictions. This retrodictive check allows model output to be compared to observations. 
The workflow encourages a focus on the full model, and replaces parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting. Any parameter (such as a trend estimate) must be carefully interpreted alongside others. %such as the  global mean or intercept and the estimate variance
Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.

@@

% More details, what's special! Feedbacks!
%emw6Apr: I think there's more than just degeneracy here, often people build in more complexity to their simulated data than their model allows, but this means they should change the model (or the question). This needs to be clearer above and below. I think this is the first point (since it's a little more obvious to readers and it relates to trends, which we always handle first), and the next one could be about structural degeneracies  but I think we need to be careful here in what we mean and explain it more to readers who will not get it. 
A key feature of this workflow is the central role of data simulation, which introduces two feedback loops. The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt a reconsideration of the model, or even a reformulation of the research question.
Further, % structural degeneracies in the model might become apparent, 
this step might reveal that some parameters are highly non-identifiable,
flagging % not sure about this word
the need to change the model structure---before incorporating real observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies may indicate a missing key driver---perhaps an expected outcome if we known our initial model was too simplistic. We can refine the model to integrate the missing process and restart the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g. unbalanced data). This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.

@@

Ecologists today have amassed data on populations and species across the globe; they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction of population and biodiversity metrics \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}. While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we believe these debates would be much reduced through use of an improved workflow. % Further, an improved workflow could give more rapid and coherent estimates, which could make it easier for  policy-makers to develop and advance initiatives aimed at slowing declines.
% Apparently conflicting results have made it harder for policy-makers to advance initiatives aimed at slowing declines \citep{gonzalez2016estimating}. While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we argue much of the work underlying these debates stems from a poor workflow.

An improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.  Using the workflow would make what now appear as major discrepancies more obviously shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process. Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected. But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model. We expect retrodictive checks on most published trend analyses would highlight major missing components in these models, and drive changes both in the models themselves and in the simulated data to check the models. While ecologists have started to use simulated data more to understand potential limitations of their models and data combined, this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility.
% expand on 'highlight major missing components in these models'? ADD example?
% Simulating data to verify and test models in trend analyses is currently rarely reported, which we believe makes it easier for ecologists to fit poor models and in turn debate them.. 
% Mention elephants rebounding...

% Trends - improvements!
Applying the workflow to current trend estimates could importantly highlight the best way to improve data collection for more reliable estimates. Returning to the example of a global estimate of trends in vertebrate populations of species over time and applying our proposed workflow would mean more efforts to define the goal and question---is it a simple global estimate? Or a need to also find which species are declining most, including those that may have poor or no data? From there a generative model using simulated data for testing could incorporate many aspects of the populations, and data, that are often only included in `null' or 'synthetic data generation' currently \citep{Buschke2021,mcrae2025utility} but could be built into the models fit to the empirical data. Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model. For example, certain populations are recovering for very specific reasons (e.g., elephants in regions where the ivory trade drove declines in the past) that perhaps should be modeled. From this model, what data are most critically needed to address the updated aims would become clearer and could drive new data collection \citep{toszogyova2024mathematical}.

@@

Increasing model complexity can be beneficial, especially when it reduces uncertainty; however, this is not always the outcome. In climate modeling, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong disagreements exist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017}.
These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. If a model becomes too complex, understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}.

Applying the workflow to process-based models would help open black box. Each successive step of model development in the workflow may highlight current problems and a path to solutions. Incorporating data simulation would introduce a crucial step between model building and data fitting, ensuring a clear delineation between the two and exposing potentially large degeneracies in the model design. Uncovering these degeneracies would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity. The workflow may also highlight degeneracies by requiring more explicit model calibration, which is currently hidden within opaque `model building,' making it easy to hide non-identifiability in the model (where certain parameters cannot be uniquely estimated). Through the workflow non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity). The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating it through projections.

Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them. The workflow could refocus attention on the research question, highlighting the ecological hypotheses that justify the use and design of the model. It would thus define a clear and limited context in which the model should apply, without always arguing about the necessity of adding more and more complexity.  % or something like : "rather than driving a constant push for increasing complexity" --- my idea is that people are always like : "oww, your model is nice, but you should add [whatever they are working on]" (classical example I heard few times: mycorrhiza! disease! pathogens!)... or "rather than constantly expanding models by adding ever more processes"? 
Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty. Instead, it would guide a step-by-step model fitting, parameter estimation, and uncertainty quantification.

@@

% Trends - outline current problem
%  Evidence of declining populations of vertebrate species in the 20th century gave rise a new subfield within ecology---conservation biology---now an entire discipline of its own that drives much of the policy-relevant science in ecology. 
% Evidence of declining populations of vertebrate species in the latter half of the 20th century, alongside increasing ecosystem health concerns, led to growing public concern about protecting and maintaining the environment, challenging ecology to help predict and prevent further losses \citep{soule1999conserving,soule1991conservation}. The idea that important taxa were declining was clear from data from certain species and their populations, such as elephants and rhinos \citep{soule1979benign,leader1990illegal}. 
% Such trends drove a number of new subfields within ecology---some of which are now complete disciplines within themselves \citep[such as conservation biology,][]{soule1985conservation}---focused on these problems, and potential ecological solutions to them CITES. Yet, as the magnitude and number of threats to these and other species have increased, with rates of habitat loss, overharvesting, pollution only increase, and anthropogenic climate change now clearly driving species loss \citep{waller2017bramble}, so has the data and its complexity. 
% Ecologists have now amassed data on populations and species over the globe, they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}. 
In the current workflow for estimating trends over time a new model with a new estimate often leads to a paper (see Fig.) because ecologists spend far too little time interrogating their models with simulated data, or their model performance fit to empirical data. 
The Living Planet Index (LPI), which aims to include long-term data on vertebrate populations of species across the globe is emblematic of these conflicting results. With updated data released semi-annually (??) alongside new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach. For example, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed most previous results \citep{Johnson2024}.

@@

\noindent % Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. 
% They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted. Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions. 
Model development is the central step of the process-based workflow, typically requiring several years, yet it often remains opaque from an external perspective. The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred. Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes. Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e. fix to a value found in the literature) or tuned to reproduce some observations or theory. The way models are currently calibrated is not a coincidence,
but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another.

@@

\centerline{\bf A common workflow to bridge trend estimation and forecasting}
\vspace*{-3mm}
{\begin{multicols}{2}
\begin{minipage}[t]{\linewidth}
\includegraphics[width=\linewidth]{figures/iterativeworkflow_details}

>>>>

\title{Closing the gap between statistical and scientific workflows for improved forecasts in ecology } 
\date{\today}
\author{Victor Van der Meersch, J. Regetz, T. J. Davies \& EM Wolkovich}
\maketitle
% $^*$ Says he is happy to help and give friendly review, but not sure he will reach level of co-author (though Lizzie used his LPI example).

@@

\begin{abstract}
Increasing biodiversity loss and climate change have led to greater demands for useful ecological models and forecasts. Relevant datasets to meet these demands have also increased in size and complexity, including in their geographical, temporal and phylogenetic scales. While new research often suggests that accounting for these complexities has a major impact on projected trends and forecasts, we argue that the typical approach to model fitting in ecology makes it impossible to evaluate and compare the models used to generate these insights and predictions. % jdavies15april: I feel this could be a critical step somehow - sharing simulations could allow cross-validation between models or perhaps cross-calibration ...
These problems stem in part from continuing gaps between statistical workflows---where the data processing and model development are often addressed separately from the ecological question and aim---and scientific workflows, where all steps are integrated. Yet, as ecologists become increasingly computational the opportunity to close this gap has never been greater. We outline how increased data simulation at multiple steps in the scientific workflow could revolutionize our understanding of ecological systems, yielding new insights for both trend estimation and forecasting.
%Combining these changes with more open model and data sharing---and developing new efforts to race the same data---could be transformative for ecological forecasting. 
A shift toward universal training in a more robust model building could bridge the gap between process-based and statistical approaches and be transformative for ecological modeling.
\end{abstract}

@@

Anthropogenic drivers are reshaping natural systems \citep{Diaz2019}. Impacts are projected to increase in coming decades, as climate change accelerates biodiversity loss, altering ecosystem services and human well-being \citep{IPBES2019}. % jdavies15april: not sure what models you are referring to here, but perhaps look up models of 'mean species abundance'.
Implementing sustainable policies to mitigate these impacts is thus a global priority, but designing the best policies requires estimating and understanding biodiversity and ecosystem trends to date alongside the skill to forecast future dynamics. % These requirements in turn have created a growing need for more global ecological data, and the types of models that can robustly handle such data.

@@

Workflows that fully integrate all the steps required to build a model from an ecological question, evaluate its limitations and degeneracies, before estimating its parameters and making projections, could reduce many of these problems. In particular, we argue that workflows that incorporate data simulation at multiple steps can quickly identify flaws in model structure and constraints in data, and allow us to undertand when, where and why different models diverge. Towards this aim,  we outline the steps of a universal workflow that could harmonize both trend estimation and forecasting.

@@

Quantitative science relies on a model-based framework, to confront hypothesis with data \citep{Chamberlin:1965cd}. In an idealized scientific method, we would formulate a research question and hypotheses, design an experiment accordingly, build a model, collect data, and using this data to inform our model and differentiate between hypotheses. % jdavies15april: could refer to the 'pre-registration' movement here ...
But this idealized method often does not apply to the reality of ecological research, and most of macroecological insights emerge from exploring patterns in data. Many important questions cannot be addressed through controlled experiments and replications. Instead, we must often rely on existing, heterogeneous datasets alongside uncertain and incomplete theory to provide a large-scale and long-term perspective \citep{Hilborn1997}.

This reality should drive researchers to use more robust and coherent methods, but the current workflows combined with the challenges ecologists are facing instead may lead to persistent problems. % jdavies15april: what challenges? forecasting from imperfect data or the urgent need to generate forecasts?
Trend estimation has focused mostly on fitting a model to empirical data in a uni-directional way % jdavies15april: not obvious what this means
that makes feedbacks to highlight uncertainty and related limitations in the model and/or data rare (Fig. \ref{fig:modeldata}). For forecasting, researchers have focused on making predictions with increasingly complex models, frequently obscuring the steps underlying model building and parameterization (Fig. \ref{fig:modeldata}). Today, researchers often calibrate different parts of the model separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole. % jdavies15april: is this particularly specific to process based models?
Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow. 
% mention the disconnection between trends and forecasts?

\begin{figure}[h]
	\centering
	\includegraphics{figures/modeldataspaces}
	\caption{Here the caption could explain a bit more the pre-model pre-data to the post-model post-data...} 
	\label{fig:modeldata}
\end{figure}
% jdavies15april: Could the bottom right be where exploratory data analysis fits in, generating hypotheses that inform the research question?

We argue a workflow that moves along the data-model space in a coherent sequence of steps with repeated data simulation % jregetz16april: Maybe “data simulation” needs to appear in the figure? Or wait to mention simulation until referencing Fig 2?
(Fig. \ref{fig:modeldata}c) could reduce many of these problems and thus improve ecological science.
The first step of this workflow is to define an explicit research question and formulate hypotheses (Fig. \ref{fig:workflowy}). This involves making clear assumptions about the most influential drivers, within the specific context of our study. This step should guide the construction of a narrative model of how we believe the system works, focusing on the mechanisms that could generate the data we observe, including the observational error. % vvdm16april: to answer Jonathan comment about verbal model
From this narrative, we can then develop a mathematical model---an ensemble of equations that encapsulates our knowledge and is designed to answer our research question. The general idea is to start with a relatively simple model that we could refine later. At this stage, prioritizing biologically meaningful parameters is crucial, as it allows us to have a sense of plausible parameter values. This means choosing a model formulation where each parameter corresponds to an interpretable behavior (which sometimes requires to think about alternative parameterizations).

With a model in place the next step focuses on testing and understanding it via data simulation. `Fake' or `test' data are generated directly from the model by fixing parameters to some reasonable range of values (which is straightforward if the parameters are interpretable) and from fake predictor data (Fig. \ref{fig:workflowy}). This simulated data should reflect the full model assumptions, and could begin to include complexities in our data structure and biases, which may in turn lead to adjusting the model. % jdavies15april: I think you could introduce this as a separate step as it seems there is a tendency for people to want to do this immediately (i.e. simulate fake data that looks like their data - I am looking at Max and Isidora ...), but I don't think this is helpful when first structuring a model.
For example, if a researcher realizes their empirical data is geographically biased, then this bias should be built into the model and thus then into this data simulation step. We then fit our model to this simulated dataset and evaluate its ability to recover the prescribed values.

Once we are confident about our model structure, we can introduce real data as part of an initial model fitting step. This way, we obtain parameter estimates constrained by observations. Here, difficulty in fitting the model might indicate an inherent need for more data to address our initial question, or re-evaluate the model of how we think the system works. This could lead us to either simplify our research question or---ideally---launch new data collection efforts. This leads to the second data simulation step, this time using our fitted model parameters to generate predictions. This retrodictive check allows model output to be compared to observations. % jregetz16april: do we need to define “retrodictive”?
The workflow encourages a focus on the full model, and replaces parameters at the core of the modeling process, as fundamental components that shape both inference and forecasting. % jregetz16april: Not sure what this means? jdavies did not like it either! (he thought it was a typo)
Any parameter (such as a trend estimate) must be carefully interpreted alongside others. %such as the  global mean or intercept and the estimate variance
Within such a workflow, forecasting emerges as a natural outcome: rather than being a final goal, it only involves jointly modeling new circumstances along with the original data.

@@

% More details, what's special! Feedbacks!
A key feature of this workflow is the central role of data simulation, which introduces two feedback loops. The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt a reconsideration of the model, or even a reformulation of the research question.
Further, % structural degeneracies in the model might become apparent, 
this step might reveal that some parameters are highly non-identifiable,
flagging % not sure about this word
the need to change the model structure---before incorporating real observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies may indicate a missing key driver---perhaps an expected outcome if we known our initial model was too simplistic. We can refine the model to integrate the missing process and restart the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure or observational biases (e.g. unbalanced data). This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes.

@@

%jregetz16april: One thing I’m stuck on (and could’ve written this note earlier too). Is “trends” really essential? Or is this just shorthand for “simple, one/few parameter, non-mechanistic forecasting models”? If yes, can we generalize that in the MS overall, or at least say so up front and then say e.g. “we’ll refer to these as trend models for brevity”? If no, and something about trend estimates is specifically key to the paper, can we say more about *that*?

Ecologists today have amassed data on populations and species across the globe; they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction of population and biodiversity metrics \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}. While shifting estimates are part of the process of science---refining our approaches and thus estimates over time---we believe these debates would be much reduced and more rapidly resolved through use of an improved workflow. % Further, an improved workflow could give more rapid and coherent estimates, which could make it easier for  policy-makers to develop and advance initiatives aimed at slowing declines.

An improved workflow that required data simulation and retrodictive checks would lead to larger model advances and a greater recognition of uncertainty---thus highlighting likely consistency in estimates across models---that could better aid policy.  Using the workflow would make what now appear as major discrepancies more obviously shifts in point estimates that are generally all in the same uncertainty space \citep{Johnson2024}---and it would challenge modelers to show major predictive advances, which is not currently part of the process. Explanatory power in most models of observational data is usually very low \citep{low2014rising,moller2002much} and thus tests of models' predictions rarely expected. But the workflow highlights that predictions from the model---what we call retrodictive checks---are part of the process of science, and critical to testing for what may be missing in a model. We expect retrodictive checks on most published trend analyses would highlight major missing components in these models, and drive changes both in the models themselves and in the simulated data to check the models. While ecologists have started to use simulated data more to understand potential limitations of their models and data combined, this is still extremely rare, and efforts to date often treat simulations as separate from the statistical model \citep{Buschke2021,dove2023quantifying}, short-circuiting their full utility.
% jdavies15april: but there is a history of using null models (e.g. Gotelli &co) that is perhaps not so different - here the 'simulated' data is some sampled process of the observed data, but we can estimate standard effect sizes that are part of our statistical model. I know this is different to the simulations you propose based on some generative process, but ecologists should be able to able to make this additional step.

% expand on 'highlight major missing components in these models'? ADD example?
% Simulating data to verify and test models in trend analyses is currently rarely reported, which we believe makes it easier for ecologists to fit poor models and in turn debate them.. 
% Mention elephants rebounding...

% Trends - improvements!
Applying the workflow to current trend estimates could importantly highlight the best way to improve data collection for more reliable estimates. Returning to the example of a global estimate of trends in vertebrate populations of species over time and applying our proposed workflow would mean more efforts to define the goal and question---is it a simple global estimate? % jdavies15april: This is what the LPI wants to be (like a index fund in the stock market tracking trends over time) , but this is often overlooked.
Or a need to also find which species are declining most, including those that may have poor or no data? From there a generative model using simulated data for testing could incorporate many aspects of the populations, and data, that are often only included in `null' or 'synthetic data generation' currently \citep{Buschke2021,mcrae2025utility} but could be built into the models fit to the empirical data. Eventually fitting the empirical data and performing retrodictive checks would likely highlight major missing components of the generative model and, ultimately, this would help inform our global estimates of mean trends. For example, certain populations are recovering for very specific reasons (e.g., elephants in regions where the ivory trade drove declines in the past) that perhaps should be modeled. From this model, what data are most critically needed to address the updated aims would become clearer and could drive new data collection \citep{toszogyova2024mathematical}.

@@

Increasing model complexity can be beneficial, especially when it reduces uncertainty; however, this is not always the outcome. In climate modeling, the uncertainty range on the effect of increasing CO$_{2}$ concentration on temperature have remained largely unchanged \citep{Zelinka2020}. This has driven calls for more rigorous and transparent calibration processes \citep{balaji2022general}. Similar concerns arise in ecology, where strong disagreements exist about the effect of climate change on future species distributions \citep{Cheaib2012} and ecosystem dynamics \citep{Lovenduski2017}.
These uncertainties have large implications beyond ecology, as they influence simulations of biosphere-atmosphere interactions and, ultimately, future climate projections \citep{Bonan2018, simpson2025confronting}.
Some researchers now advocate for the simplification of models, to avoid over-parametrization when the data provide little information to constrain some parameters \citep{Wang2017, Harrison2021}. % jdavies15april: need to distinguish this from multiple model comparisons that attempt to identify the 'minimum adequate model' - which lead to overfitting.
If a model becomes too complex, understanding the sources of uncertainty and how they propagate through the model may become nearly impossible.
Each additional process and parameter can increase overall uncertainty to the point where model projections lose their usefulness for decision makers \citep{Saltelli2020}. % jdavies15april: complexity, such as interaction terms also requires more data to fit, and thus adding parameters can increase parameter uncertainty ....

Applying the workflow to process-based models would help open the black box. % jdavies15april: what is the black box?
Each successive step of model development in the workflow may highlight current problems and a path to solutions. Incorporating data simulation would introduce a crucial step between model building and data fitting, ensuring a clear delineation between the two and help expose potential degeneracies in the model design. %jregetz16april: Maybe good to define the term degeneracy here (like we do for non- identifiability below)
Uncovering degeneracies would likely force researchers to begin with a simpler version of the model, which they could build on iteratively, testing for support---or lack thereof---when adding model complexity. The workflow may also highlight degeneracies by requiring more explicit model calibration, % jdavies15april: still not really sure what is 'model calibration' and how it differs from model fitting and parameter estimation
which is currently hidden within opaque `model building', making it easy to hide non-identifiability in the model (where certain parameters cannot be uniquely estimated). Through the workflow non-identifiable parts of models could be addressed by reformulating the mathematical structure of certain processes, or finding ways to apply additional constraints (e.g., narrower ranges for certain parameters or developing new hypotheses that target the appropriate level of complexity). The resulting process-based models would likely be simpler and thus more tractable for quantifying parameter uncertainty and propagating it through projections. 
% jdavies15april: it would be nice to also have a clear empirical example (like the LPI above) in this section - are we projecting species distributions, ecosystem properties? This could help illustrate potential degeneracies etc.

% jdavies15april: might be better grouped together in concluding section
Beyond improving the model building, the workflow also has the potential to shift how process-based models are perceived, particularly by those unfamiliar with them. The workflow could refocus attention on the research question, highlighting the ecological hypotheses that justify the use and design of the model. It would thus define a clear and limited context in which the model should apply, without always arguing about the necessity of adding more and more complexity.  % or something like : "rather than driving a constant push for increasing complexity" --- my idea is that people are always like : "oww, your model is nice, but you should add [whatever they are working on]" (classical example I heard few times: mycorrhiza! disease! pathogens!)... or "rather than constantly expanding models by adding ever more processes"? 
Process-based model would once again be a way to answer a research question---whereas today, model simulations have increasingly become a subject of study on their own.
Ideally, applying the workflow would help to move away from the traditional process-based model paradigm, where parameters are typically assigned fixed values without properly accounting for their uncertainty. Instead, it would guide a step-by-step model fitting, parameter estimation, and uncertainty quantification.

@@

% Trends - outline current problem
%  Evidence of declining populations of vertebrate species in the 20th century gave rise a new subfield within ecology---conservation biology---now an entire discipline of its own that drives much of the policy-relevant science in ecology. 
% Evidence of declining populations of vertebrate species in the latter half of the 20th century, alongside increasing ecosystem health concerns, led to growing public concern about protecting and maintaining the environment, challenging ecology to help predict and prevent further losses \citep{soule1999conserving,soule1991conservation}. The idea that important taxa were declining was clear from data from certain species and their populations, such as elephants and rhinos \citep{soule1979benign,leader1990illegal}. 
% Such trends drove a number of new subfields within ecology---some of which are now complete disciplines within themselves \citep[such as conservation biology,][]{soule1985conservation}---focused on these problems, and potential ecological solutions to them CITES. Yet, as the magnitude and number of threats to these and other species have increased, with rates of habitat loss, overharvesting, pollution only increase, and anthropogenic climate change now clearly driving species loss \citep{waller2017bramble}, so has the data and its complexity. 
% Ecologists have now amassed data on populations and species over the globe, they have also engaged in an increasing number of debates on regional and global trends, with arguments over the magnitude and even direction \citep{Dornelas2014,Leung2020,terry2022no,muller2024weather}. 
In the current workflow for estimating trends over time a new model with a new estimate often leads to a paper (see Fig.) because ecologists spend far too little time interrogating their models with simulated data, or their model performance fit to empirical data. 
The Living Planet Index (LPI), which aims to include long-term data on vertebrate populations of species across the globe is emblematic of these conflicting results. % jdavies16april: could highlight how influential the LPI has become (recent reports suggest a 73% ave. pop decline since 1970. This would be massive!
With updated data released semi-annually (??) alongside new estimates of decline, a growing number of high-profile papers have challenged how strong the evidence is for population decline \citep{Dornelas2014,gonzalez2016estimating,wagner2021insect,muller2024weather}, with each paper taking a slightly different analytical approach. For example, \citet{Leung2020} published a mixture model that suggested most populations were not significantly declining, followed by other alternative modeling approaches \citep{Buschke2021,puurtinen2022living} including a recent one suggesting a basic analysis of the dataset should always include three sources of autocorrelation, finding trends that encompassed most previous results \citep{Johnson2024}. 
% jdavies16april: One of my major issues with all these studies is that they have lost track of the purpose of the index, which is to capture general trends in animal populations. We do not actually care (much) about the population trends in the LPI, but how these are representative of global trends. For this we need to predict outside of the set ...

@@

\noindent % Process-based models are built on explicit mathematical equations to describe (supposedly causal) relationships between environmental drivers and ecological responses. 
% They also often incorporate empirical relationships, particularly when knowledge is incomplete or when some processes are intentionally omitted. Processes are often represented at different nested spatiotemporal scales, depending on the underlying assumptions. 
Model development is the central step of the process-based workflow, typically requiring several years, yet it often remains opaque from an external perspective. The step of designing the model---translating knowledge and hypotheses into mathematical equations and parameters---is often blurred with the step of model calibration (or tuning), where parameter values are inferred. Models are often treated as an accumulation of multiple submodels, each governing one or several ecological processes. Rather than being fitted as a whole, submodels are calibrated separately against specific subsets of data, and some parameters are simply prescribed (i.e. fix to a value found in the literature) or tuned to reproduce some observations or theory. The way models are currently calibrated is not a coincidence, but rather an inappropriate way to accommodate their complexity, where many parameters compensate for one another. % jdavies15april: is this more specific to process based  models?

@@

\centerline{\bf A common workflow to bridge trend estimation and forecasting} % jdavies15april: would it be possible to label this panel with components of the LPI to provide a clearer illustration of how the framework is implemented? % vvdm16april: not sure I understood
\vspace*{-3mm}
{\begin{multicols}{2}
\begin{minipage}[t]{\linewidth}
\includegraphics[width=\linewidth]{figures/iterativeworkflow_details}
